---
title: "Normalization"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dependencies

Load the major packages needed for this analysis first. Additional packages will be loaded as needed.

```{r, message=FALSE, warning=FALSE, echo = TRUE}

# dependencies needed
library(SingleCellExperiment)
library(scater)
library(scran)
library(uwot)
library(Rtsne)
```


Loading the filtered file which was output by "2_QC_Filtering.Rmd"
```{r, echo = TRUE}
filtered <- readRDS(file="../processed_data/guiu_filtered.rds")
```

## Removing composition bias by normalization

To perform normalization by deconvolution first we have to perform a quick clustering step.
```{r, echo=TRUE}


library(scran)
set.seed(100)
clust.guiu <- quickCluster(filtered) # first do quick clustering 
table(clust.guiu) # how many clusters and how many cells in each


```

Next, size (sum) factors are estimated from pooled counts and then "deconvolved" to obtain value for each cell.

```{r, echo=TRUE}
filtered <- computeSumFactors(filtered, cluster=clust.guiu)
summary(sizeFactors(filtered))
```

Using the above calculated size factors, compute the normalized expression values for each cell. Log-transform the normalized expression values for downstream analysis
```{r, echo=TRUE}
filtered <- logNormCounts(filtered)
assays(filtered) # now there are two assays the counts & the log normalized counts just made

```

## Modeling the variance and selection of highly variable genes (HVGs)

Compute variance of the logcounts for each gene across all cells and then rank the output based on proportion of variance that is biological rather than technical.
```{r, echo=TRUE}
set.seed(101)
filtered.var <- modelGeneVar(filtered)
filtered.var <- filtered.var[order(filtered.var$bio, decreasing=TRUE),]
head(filtered.var)
```

Select the most highly variable genes either by strict 10% proportion or by a 1% FDR threshold. See how many genes are retained by each metric.

```{r, echo=TRUE}
top.genes.var <- getTopHVGs(filtered.var, prop=0.1) # keeps the top 10% as HGVs
length(top.genes.var) 

top.genes.varFDR <- getTopHVGs(filtered.var, fdr.threshold = 0.01) 
length(top.genes.varFDR) 

```
Plot log normalized expression of genes with the largest biological component to their variance
```{r, echo=TRUE}
plotExpression(filtered, features=top.genes.varFDR[1:10])
```

Save out the resulting files

```{r, echo=TRUE}

saveRDS(filtered, file="../processed_data/guiu_filtered_normalized.rds") # save the sce post norm
saveRDS(top.genes.varFDR, file="../processed_data/top.genes.varFDR.rds") 
saveRDS(filtered.var, file="../processed_data/filtered.var.rds") 
```


```{r, echo=TRUE}
sessionInfo()
```
